\section{Conclusion}
\label{sec:conclusion}

We have presented \sozkz{}, a family of small language models trained
from scratch exclusively on Kazakh text at four parameter scales
(50M, 150M, 300M, and 600M).  Our results demonstrate that dedicated
monolingual models, paired with a purpose-built tokenizer, can achieve
competitive performance on core Kazakh NLP tasks while requiring
orders-of-magnitude fewer parameters than multilingual alternatives.
The \sozkz{}-600M model achieves a BPB of \bpbSozKZSixHundredM{} on
held-out Kazakh text, substantially outperforming all multilingual
models including those with 10$\times$ or more parameters, and reaches
\accSentimentSozKZSixHundredM{} on sentiment classification---competitive
with models many times its size.

\paragraph{Limitations.}
Our models exhibit clear limitations on knowledge-intensive tasks.
On multiple-choice question answering, even the largest \sozkz{} model
scores \accMcQaSozKZSixHundredM{}, well below the performance of
7B+ multilingual models that benefit from extensive world knowledge
absorbed during training on much larger corpora.  Similarly, on
Belebele reading comprehension, the gap between our 600M model
(\accBelebeleSozKZSixHundredM{}) and larger multilingual models
remains substantial.  These results underscore that model
specialization provides efficiency gains on language-centric tasks
but cannot substitute for the parametric knowledge stored in larger
models.  Additionally, our training data is limited to publicly
available Kazakh web text, which constrains both the domain coverage
and the factual knowledge our models can acquire.

\paragraph{Future Work.}
Several directions merit further investigation.  First, scaling
beyond 600M parameters---particularly to the 1--3B range---would test
whether the efficiency advantages of dedicated models persist at
larger scales.  Second, instruction tuning and alignment of the base
models could unlock practical applications in dialogue and
task-following scenarios.  Third, fine-tuning on downstream tasks such
as named entity recognition and question answering may close the gap
with larger models on knowledge-intensive benchmarks.  Finally,
extending our approach to other Turkic languages (Turkish, Uzbek,
Kyrgyz) could leverage shared morphological structure through
multilingual dedicated models at modest scale.

We release all \sozkz{} models, the tokenizer, and the training
pipeline under open licenses\footnote{Models available at
\url{https://huggingface.co/stukenov}} to support further research
on efficient language modeling for underserved languages.
