\section{Experiments}
\label{sec:experiments}

We evaluate the \sozkz{} models and a set of multilingual competitors on
\numTasks{}~benchmarks spanning language modeling quality, reading comprehension,
question answering, sentiment analysis, named entity recognition, and topic
classification. All evaluations are conducted in a zero-shot setting.

% ---------------------------------------------------------------------------
\subsection{Benchmarks}
\label{sec:benchmarks}

\paragraph{Bits-per-Byte (BPB).}
We measure language modeling quality using bits-per-byte on held-out Kazakh text
(FLORES-200 \texttt{kaz\_Cyrl} and Kazakh Wikipedia). BPB is computed as
$\text{BPB} = \text{NLL} / (N_{\text{bytes}} \cdot \ln 2)$, where
$N_{\text{bytes}}$ counts UTF-8 bytes. Since Kazakh Cyrillic characters are
encoded as 2~bytes each, this metric normalizes across tokenizers with different
vocabularies. Lower is better.

\paragraph{MC QA.}
Multiple-choice question answering on the \texttt{kk-socio-cultural-bench-mc}
dataset. Each question has 4~answer choices. We score using full answer text
likelihood---the sum of log-probabilities over all tokens in each candidate
answer, length-normalized---rather than single-token logit comparison, which
would be biased by tokenizer vocabulary alignment. Accuracy is reported.

\paragraph{Sentiment.}
Kazakh sentiment classification using logit-based scoring with Kazakh-language
target words as class labels. This avoids generation-based evaluation and
provides a consistent comparison across models with different vocabularies.
Accuracy is reported.

\paragraph{Belebele.}
Reading comprehension from the Belebele benchmark~\citep{bandarkar2023belebele}
(\texttt{kaz\_Cyrl} subset). Each item presents a passage and a question with
4~multiple-choice answers. Scoring uses full answer text likelihood as in MC~QA.
Accuracy is reported.

\paragraph{NER.}
Named entity recognition on KazNERD~\citep{yeshpanov2022kaznerd}. The original
25~fine-grained entity classes are collapsed into 6~simplified categories
(Person, Location, Organization, Date, Number, Miscellaneous) for tractable
zero-shot scoring. F1 is reported.

\paragraph{SIB-200.}
Topic classification on the Kazakh subset of the SIB-200
benchmark~\citep{adelani2024sib200}. Each text is assigned to one of 7~topic
categories. Accuracy is reported.

% ---------------------------------------------------------------------------
\subsection{Models}
\label{sec:models}

We evaluate \numModels{}~models organized into two groups.

\paragraph{SozKZ family (ours).}
Four Kazakh-only models trained from scratch on ${\sim}9.0$B~Kazakh tokens,
as described in \Cref{sec:methodology}: \sozkz{}-50M, \sozkz{}-150M,
\sozkz{}-300M, and \sozkz{}-600M. These models range from 50M to 587M
parameters and use a dedicated Kazakh BPE tokenizer.

\paragraph{Multilingual competitors.}
Ten general-purpose multilingual models that include Kazakh in their training
mixture but are not specialized for it:
\begin{itemize}
  \item Qwen-2.5 (0.5B, 1.5B, 7B)~\citep{bai2023qwen},
  \item Llama-3 (1B, 3B, 8B)~\citep{touvron2023llama2},
  \item Gemma-2 (2B, 9B)~\citep{team2024gemma},
  \item Mistral-7B~\citep{jiang2023mistral},
  \item GPT-OSS-120B (largest open-source baseline).
\end{itemize}
All competitor models are evaluated using their native tokenizers and default
configurations. The comparison is deliberately asymmetric: our smallest model
(50M) is 10$\times$ smaller than the smallest multilingual competitor (Qwen-0.5B),
while our largest (600M) is 12$\times$ smaller than Llama-3-8B.

% ---------------------------------------------------------------------------
\subsection{Evaluation Protocol}
\label{sec:protocol}

All evaluations use the same pipeline (\texttt{scripts/eval/}) to ensure
consistency. Key protocol details:

\begin{itemize}
  \item \textbf{Logit-based scoring} for all classification and multiple-choice
        tasks. No text generation is required, which eliminates sensitivity to
        decoding hyperparameters.
  \item \textbf{Full answer likelihood} for multiple-choice tasks: each
        candidate answer is scored by the sum of its token log-probabilities
        conditioned on the question, normalized by token count.
  \item \textbf{Zero-shot} evaluation for all tasks and models---no in-context
        examples or task-specific fine-tuning.
  \item \textbf{Consistent hardware}: efficiency comparisons (latency,
        throughput) use the same GPU (NVIDIA A10) for all models, with median
        latency reported rather than mean to mitigate outlier sensitivity.
\end{itemize}
