\section{Introduction}

Kazakh is a Turkic language spoken by approximately 22 million people,
primarily in Kazakhstan and neighboring countries.  As an agglutinative
language with rich morphology, Kazakh poses particular challenges for
natural language processing: words carry extensive inflectional and
derivational suffixes, leading to a large effective vocabulary that
general-purpose multilingual tokenizers handle poorly.  Despite a
growing body of NLP research for Kazakh~\citep{yeshpanov2022kaznerd,
yeshpanov2024kazqad}, the language remains underserved by the current
generation of large language models, which typically allocate only a
fraction of their capacity to low-resource languages.

Modern multilingual models such as Llama~3~\citep{touvron2023llama},
Gemma~\citep{team2024gemma}, Qwen~2.5~\citep{bai2023qwen}, and
Mistral~\citep{jiang2023mistral} are trained on corpora dominated by
English and other high-resource languages.  When applied to Kazakh,
these models face two compounding inefficiencies.  First, their
tokenizers fragment Kazakh text into disproportionately many subword
tokens---a phenomenon quantified by \emph{fertility}, the average
number of tokens per word.  Second, the vast majority of model
parameters encode knowledge about languages other than Kazakh, yielding
a poor ratio of useful capacity to total model size.

We argue that \textbf{small, dedicated models can match or exceed much
larger multilingual models on Kazakh language tasks with
orders-of-magnitude better efficiency}.  To test this thesis, we train
\sozkz{}, a family of Llama-architecture causal language models at four
parameter scales---50M, 150M, 300M, and 600M---entirely from scratch on
a curated Kazakh corpus of approximately 9 billion tokens.

Our contributions are as follows:
\begin{enumerate}
  \item We present an end-to-end pipeline for training Kazakh language
    models from scratch at four scales (50M--600M parameters), including
    data collection, cleaning, and pre-tokenization.
  \item We design a dedicated ByteLevel BPE tokenizer with a 50K
    vocabulary\footnote{\url{https://huggingface.co/stukenov/kazakh-bpe-32k}}
    that achieves a 2--3$\times$ fertility advantage over
    multilingual tokenizers on Kazakh text.
  \item We construct a \numTasks{}-task benchmark suite and evaluate
    \sozkz{} models against \numModels{} competitors spanning
    0.5B--120B parameters, providing the most comprehensive comparison
    of language model performance on Kazakh to date.
  \item We report empirical scaling curves for Kazakh language modeling,
    identifying the efficiency sweet spot where dedicated small models
    offer the best performance-per-parameter trade-off.
  \item We release all models, the tokenizer, and the training pipeline
    under open licenses to support further research on Kazakh NLP.
\end{enumerate}

The rest of this paper is organized as follows.
\Cref{sec:related} surveys related work on Kazakh NLP, low-resource
language modeling, scaling laws, and tokenizer design.
\Cref{sec:methodology} describes our data pipeline, tokenizer,
model architectures, and training procedure.
\Cref{sec:experiments} details the benchmark suite and evaluation
protocol.  \Cref{sec:results} presents and analyzes results.
\Cref{sec:conclusion} summarizes our findings and outlines future
directions.
