\section{Related Work}
\label{sec:related}

\subsection{Kazakh NLP}

Research on Kazakh natural language processing has accelerated in
recent years, driven largely by the IS2AI research group at Nazarbayev
University.  \citet{yeshpanov2022kaznerd} introduced KazNERD, a
large-scale named entity recognition dataset for Kazakh covering 25
entity classes across news, fiction, and legal domains.
\citet{yeshpanov2024kazqad} subsequently released KazQAD, a
question-answering dataset for Kazakh built from Wikipedia and
educational texts.  These resources have enabled systematic evaluation
of NLP models on Kazakh for the first time.

Beyond datasets, several efforts have targeted Kazakh language
modeling directly.  Multilingual models such as
mBERT~\citep{devlin2019bert} and XLM-R~\citep{conneau2020xlmr}
include Kazakh in their training data but allocate minimal capacity to
it.  More recently, KazLLM~\citep{kazllm2024} explored continued
pre-training of large models on Kazakh corpora.  Our work differs in
training dedicated models entirely from scratch, allowing full control
over architecture, tokenizer, and data composition.

\subsection{Low-Resource Language Models}

Training language models for low-resource languages has emerged as an
active research area.  \citet{ogueji2021afriberta} demonstrated with
AfriBERTa that small transformer models trained on modest corpora
(less than 1\,GB) can achieve competitive performance on African
languages, challenging the assumption that large-scale data is a
prerequisite.  For Arabic, \citet{sengupta2023jais} trained Jais, a
13B-parameter model from scratch on a bilingual Arabic-English corpus,
showing that dedicated models outperform multilingual ones on
Arabic-specific benchmarks.

SEA-LION~\citep{sealion2023} adopted a similar philosophy for
Southeast Asian languages, training models at multiple scales with
language-specific tokenizers.  The BLOOM
project~\citep{bigscience2022bloom} took a different approach,
training a single 176B-parameter multilingual model on 46 languages
with the goal of equitable coverage.  Our work follows the
dedicated-model approach of Jais and AfriBERTa but at substantially
smaller scales (50M--600M parameters), testing how far efficiency
gains from specialization can compensate for reduced model capacity.

\subsection{Scaling Laws}

\citet{kaplan2020scaling} established power-law relationships between
model size, dataset size, compute budget, and cross-entropy loss for
neural language models.  \citet{hoffmann2022chinchilla} refined these
findings with the Chinchilla scaling laws, demonstrating that many
large models are significantly undertrained relative to their parameter
count and that a compute-optimal model should be trained on
approximately 20 tokens per parameter.

In practice, the Llama model family~\citep{touvron2023llama,
touvron2023llama2} demonstrated that training smaller models on
substantially more data than the Chinchilla-optimal ratio yields
models that are more efficient at inference time.  Our training regime
follows this over-training philosophy: the \sozkz{}-600M model is
trained on approximately 9B tokens, yielding a tokens-to-parameters
ratio of roughly 15:1---slightly below the Chinchilla optimum but
reflective of the available Kazakh data.

These scaling laws have been derived primarily from English-language
experiments.  Whether they transfer to morphologically rich,
low-resource languages with different tokenizer efficiencies remains
an open question that our scaling analysis addresses.

\subsection{Tokenizer Design}

Subword tokenization, introduced by \citet{sennrich2016bpe} with
Byte-Pair Encoding (BPE) and refined by
\citet{kudo2018sentencepiece} with SentencePiece, is now standard in
language model pre-training.  The choice of tokenizer has outsized
effects on model efficiency for morphologically rich
languages~\citep{rust2021tokenizer}: a tokenizer trained on
English-dominated data will fragment agglutinative languages like
Kazakh, Turkish, or Finnish into far more tokens per word than
necessary, effectively reducing the model's context window and
increasing inference cost per semantic unit.

\emph{Fertility}---the average number of tokens per whitespace-separated
word---is a widely used metric for quantifying this
inefficiency~\citep{acs2019fertility}.  Prior work on Turkic
languages has shown that dedicated tokenizers achieve fertility
values 2--3$\times$ lower than multilingual
alternatives~\citep{toraman2023turkic}.

We train a ByteLevel BPE tokenizer with a 50K vocabulary exclusively
on Kazakh text and quantify its fertility advantage over the
tokenizers used by Llama~3, Gemma, Qwen~2.5, and Mistral.  This
dedicated tokenizer is a key component of the efficiency gains we
observe, as lower fertility translates directly to more content per
fixed context window.
